A significance test in forward stepwise regression

### 2014-03-21 paper/paper.tex
Paper: some small edits

Forward stepwise regression is a classical model selection procedure
that begins with an empty model and sequentially adds the best predictor
variable in each step. Classical significance tests fail
when a model
has been selected this way and tend to be anti-conservative.
In the lasso setting, \cite{significance:lasso} derived a novel test
statistic along with a correct asymptotic distribution, making possible
valid inferences after model selection.
\cite{tests:adaptive} modified and extended those results to the
group lasso \citep{grouplasso} and other penalized regression
problems, but only under the global null hypothesis.
One of the strengths of these test statistics is that they can be
used for valid significance testing when computed on the same
data used for model selection, eliminating the need for data splitting.
The present work iteratively applies the global null test
of \cite{tests:adaptive} for each step in forward stepwise selection,
and works out some of the details necessary for
models with grouped variables. The resulting method can
be more statistically efficient than validation on held-out data, and
more computationally efficient than penalized methods with
regularization parameters chosen by cross-validation.

@@

Some classical attempts to address this issue include Monte Carlo
estimation of tables of adjusted $F$-statistic values \citep{mc:ftoenter},
and permutation statistics \citep{permutation:stop}. Aside from the works
this paper is based on, there have been other recent attempts to do
inference after model selection. Most of these make use of subsampling
\citep{meinshausen:buhlmann} or data splitting \citep{wasserman:roeder}.
Our approach allows use of the full data and does not require the
extra computation involved in subsampling.
Before describing the full approach we first introduce notation and
specify our implementation of forward stepwise, which is slightly
different from the most commonly used versions.

We allow forward stepwise selection to add groups of variables in each
step, not only in the case of binary encoding for categorical
variables but also for any grouping purpose. For example, groups of
variables may be pre-designated factors such as expression
measurements for all genes in a single functional pathway.
To emphasize this we will use $g, h$ as indices rather than the usual
$i, j$ throughout. Since single variables can be considered groups of
size 1, our general exposition includes non-grouped situations as a special
case. Our variant of forward stepwise uses a different method than
usual for choosing which group to add because we are using the
same objective in the forward stepwise procedure that was used to
derive the test statistics in penalized regression settings.

@@

Define $p = \sum_{g=1}^Gp_g$,
the total number of individual variables, so $p = G$ in the
case where all groups have size 1. Let $X$ be the matrix constructed
by column-binding the $X_g$, that is  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation*}
X = \begin{pmatrix} X_1 & X_2 & \cdots & X_G  \end{pmatrix}
\end{equation*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We assume throughout that the individual columns of $X$ are scaled to
have unit 2-norm (we do not scale groups, this is accomplished later
by weighting).
With each group we associate the $p_g \times 1$ coefficient vector
$\beta_g$, and write $\beta$ for the $p \times 1$ vector constructed
by stacking all of the $\beta_g$ in order.  Finally, our model for the
response is
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\begin{aligned}
\label{eq:gmodel}
y & = X \beta + \sigma \epsilon \\
   & = \sum_{g=1}^G X_g \beta_g + \sigma \epsilon
\end{aligned}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
where $\epsilon$ is noise. Unless otherwise specified we assume
i.i.d. Gaussian noise $\epsilon \sim N(0, I_{n \times n})$ and that
$\sigma$ is known.

We allow for the possibility $p > n$ in which case \eqref{eq:gmodel} is
generically underdetermined. In such cases it still may be possible to
estimate $\beta$ well if it is sparse--that is, if it has few nonzero
entries. In the rest of this paper we refer to variable groups $X_g$
as noise groups if $\beta_g$ is a zero vector and as true or signal
groups if $\beta_g$ has any nonzero entries. We refer to the number
of such nonzero groups as the \textit{sparsity} of the model, and
denote this $k := \# \{ g : \beta_g \neq 0 \}$.

Before we describe the forward stepwise procedure we require one last
ingredient. To each group of variables we assign a weight $w_g$. These
weights act like penalties or costs, so increasing $w_g$ makes it
more difficult for the group $X_g$ to enter the model. The
modeler can choose weights arbitrarily, but we will mostly use one
particular choice, based on $p_g$, that we discuss later. With this we
are ready to describe the forward stepwise procedure.

@@

\begin{algorithm}
  \caption{Forward stepwise variant with groups and weights}
  \label{algo:fs}
  \begin{algorithmic}[1]
    \REQUIRE An $n$ vector $y$ and $n \times p$ matrix $X$ of $G$ variable groups with weights $w_g$
    \ENSURE Ordered active set $A$ of variable groups included in the model at each step
    \STATE $A \gets \emptyset$, $A^c \gets \{ 1, \ldots, G\}$
    \FOR{$s=1$ to $steps$}
    \STATE $g^* \gets \argmax_{g \in A^c} \{ \norm{X_g^T r_{s-1}}_2 / w_g \}$
    \STATE $P \gets I - X_{g^*}X_{g^*}^\dagger$
    \STATE $A \gets A \cup \{ g^* \}$, $A^c \gets A^c \backslash \{ g^* \}$
    \FORALL{$g \in A^c$}
      \STATE $X_g \gets P X_g$
    \ENDFOR
    \STATE $r_s \gets P r_{s-1}$
    \ENDFOR
    \RETURN $A$
  \end{algorithmic}
\end{algorithm}

In the common case where all groups are single variables this version
coincides with the standard implementations of forward stepwise.
We will now use forward stepwise to refer specifically to
Algorithm~\ref{algo:fs} unless otherwise specified. Note that other
implementations of forward stepwise use different criteria for choosing
the next variable to add, such as the correlation with the residual.
Since we do not renormalize the columns after projecting the covariates
(lines 6 to 8 above), and since we have weights, we are not computing
correlations unless the design matrix is orthogonal and all weights are 1.
There are advantages and disadvantages to both criteria which we do not
discuss. Our choice was motivated by the group lasso result in
\cite{tests:adaptive}, but we believe other criteria can be handled
with appropriate modifications.

@@

\cite{significance:lasso} prove that under the null hypothesis that $A_k$
contains all the strong predictor variables, the distribution of a
test statistic $T_k \propto \lambda_k(\lambda_k - \lambda_{k+1})$ is
asymptotically Exp(1). In the lasso case we know a lot about the knots
and active set, but the group lasso picture is slightly more
complicated. For the group lasso, $\hat \beta(\lambda)$ does not have
piecewise linear components. To overcome this difficulty we will
restrict our attention to the first group of variables to enter the
active set since the analysis then follows almost exactly as for the
lasso.

@@

We now describe how to calculate our p-value, following the discussion of
group lasso in \cite{tests:adaptive}. For the rest of this section let
$g = \gstar$ be the index of
the group attaining the maximum on line 3 of Algorithm~\ref{algo:fs}.
%\lambda = \norm{X_g^Ty}_2/w_g$ be the value of the maximum, and
%$r = \text{rank}(X_g)$ be the rank of the maximizing group.
For a vector $u \in \real^p$, let $u_h$ denote the coordinates
of $u$ corresponding to the columns of group $h$ in the design matrix $X$.
We can rearrange the columns of $X$ to group these adjacently, so that
\[
u^TX^T = \begin{pmatrix} u_1^T X_1^T & u_2^T X_2^T & \cdots & u_G^TX_G^T \end{pmatrix}
\]
One step of the calculation will be to find an orthonormal basis for the
linear space $\vecsp_g = \{ u \in \real^p : u_g^T X_g^T y = 0, u_h = 0
\text{ for all } h \neq g \}$ so that we can project orthogonally to
this space. If $X_g$ is a single column and $X_g^Ty \neq 0$ (which
should be the case since $g$ maximizes the absolute value of this
quantity), then the
space is trivial and the desired orthogonal projection is the identity.
Otherwise, if $X_g$ has $p_g > 1$ columns the space $\vecsp_g$ will
generally have dimension $p_g-1$.
% have dimension $r-1$ for $r = \text{rank}(X_g)$. This holds except in
% degenerate cases such as $X_g = 0$ or $r \ll p_g$, but these cases are not
% likely to occur since $\norm{X_g^T y}_2/w_g$ is supposed to be maximized.
% We proceed by forming the $n \times p_g$ matrix
% $X_g - X_g X_g^T y y^T X_g/\norm{X_g^Ty}_2^2$ and performing Gram-Schmidt
% on the first $r-1$ columns. For precise theorems showing that this will
% generally work, see \cite{outerp:indep}.
We compute an orthonormal basis by Gram-Schmidt and form a
$p_g \times p_g$ matrix, which we denote $V_g$, by appending one column
%0's in the coordinates corresponding to all groups $h \neq g$ and to
of zeroes (since Gram-Schmidt only produces $p_g-1$ columns).
With this we are ready to define the projection

@@

\begin{algorithm}
  \caption{Computing p-value}
  \label{algo:pval}
  \begin{algorithmic}[1]
    \REQUIRE Response $y$, grouped design matrix $X$ with weights, inactive set $A^c$, index $g$ of last group to enter active set $A$.
    \ENSURE Our p-value for the group $g$ entering the model.
    \STATE Compute $H_g$ and $\sigma^2_g$
    \IF{$X_g$ is a single column}
    \STATE $\sigma_g^2 \gets X_g^T\Sigma X_g/w_g^2$
    \STATE $\tilde X_g \gets X_g / w_g \cdot \text{sign}(X_g^Ty)$
    \ELSE
    \STATE $P_g \gets \Sigma XV_g (V_g^T X^T \Sigma X V_g)^\dagger V_g^TX^T$
    \STATE $H_g \gets (I - P_g)\Sigma$
    \STATE $\sigma^2_g \gets y^TX_gX_g^T H_g X_gX_g^Ty / (w_g^2 \norm{X_g^Ty}_2^2)$
    \STATE $\tilde X_g \gets H_g X_g X_g^T y /(\norm{X_g^Ty}_2 w_g)$
    \ENDIF
    \STATE $\lambda \gets \norm{X_g^Ty}_2/w_g$, $r \gets \text{rank}(X_g)$
%    \STATE \COMMENT{Compute the following two $p$-vectors}
    \STATE $a \gets X^T(y - \lambda \tilde X_g)$
    \STATE $b \gets X^T \tilde X_g$
    \STATE Compute solution $(v_-, v_+)$ of \textsc{LinFrac}(a,b) optimization subproblem
    \STATE $u \gets [F_{\chi^2_r}(v^2_-/\sigma_g^2) - F_{\chi^2_r}(\lambda^2/\sigma_g^2)]/[F_{\chi^2_r}(v^2_-/\sigma_g^2) - F_{\chi^2_r}(v^2_+/\sigma_g^2)]$
    \RETURN $u$
  \end{algorithmic}
\end{algorithm}

We next describe the \textsc{LinFrac} subproblem and its solution. We
gave the problem this name because it originated in the form
\[
\maximize_{h \neq g, \norm{u_h}_2 = 1} \frac{u_h^TX_h^Ty - u_h^TX_h^TX_gX_g^Ty}
{1 - u_g^TX_g^TX_hu_h}
\]
The solution we describe next is to a slightly different problem, and
also incorporates the information that $g$ maximizes $\norm{X_g^Ty}_2/w_g$.
For details see Appendix~\ref{app:closedform} or \cite{tests:adaptive}.

\begin{algorithm}
  \caption{The \textsc{LinFrac} optimization subproblem}
  \label{algo:linfrac}
  \begin{algorithmic}[1]
    \REQUIRE The $p$-vectors $a, b$ in Algorithm~\ref{algo:pval}, weights, inactive set $A^c$, a small tolerance number
%(we take $1e^{-10}$)
    \ENSURE Solution pair $(v_-, v_+)$.
    \FOR{$h$ in $A^c$}
      \IF{$\norm{b_h}_2 == 0$ or $\norm{a_h}_2/\norm{b_h}_2 < \text{tol}$}
        \STATE $(v^-_h, v^+_h) \gets (0, \infty)$
      \ELSE
        \STATE $\theta_c \gets a_h^Tb_h/(\norm{a_h}_2 \norm{b_h}_2)$
        \STATE $\theta_s \gets \sqrt{1-\theta_c^2}$
        \STATE $\theta \gets \arccos (\theta_c)$
        \STATE $\phi_s \gets \theta_s \norm{b_h}_2 / w_h$
        \IF{$\phi_s > 1$}
          \STATE $(v^-_h, v^+_h) \gets (0, \infty)$
        \ELSE
          \STATE $\phi \gets \arcsin(\phi_s)$
%          \STATE $\phi_2 \gets \pi - \phi_1$
          \STATE $z_\pm \gets s_\pm \norm{a_h}_2 \cos(\phi) /(w_h - s_\pm \norm{b_h}_2 \cos(\theta - \phi))$ for $s_\pm = \pm 1$
          \IF{$\norm{b_h}_2 < w_h$}
            \STATE $(v_h^-, v_h^+) \gets (\max \{ z_+, z_- \}, \infty)$
          \ELSE
            \STATE $(v_h^-, v_h^+) \gets (\min \{ z_+, z_- \}, \max \{ z_+, z_- \})$
          \ENDIF
        \ENDIF
      \ENDIF
    \ENDFOR
    \STATE $v^- \gets \min_h v_h^-$
    \STATE $v^+ \gets \max_h v_h^+$
    \RETURN $(v^-, v^+)$
  \end{algorithmic}
\end{algorithm}

@@

\subsection{Glinternet for hierarchical interactions}
\label{sec:glint}
In regression settings with many variables, considering models with pairwise interactions can drastically increase model complexity. \cite{glint} propose a method called \textsc{glinternet} to reduce the statistical complexity of this problem. The method imposes a strong hierarchical constraint on interactions (similar to that in \cite{bien:hierarchical}) where an interaction term can only be included if both its main effects are also included. They accomplish this by creating a design matrix with both main effects alone and also with groups including main effects with their first order interactions. Then they fit a group lasso model with the expanded design matrix. Because interaction terms only appear in groups along with their respective main effects, the hierarchy condition holds for the fitted model. We now consider a related procedure as an example problem, but first modify their method to simplify some parts. Let the expanded design matrix be given by
\begin{equation}
\label{eq:glintmat}
\tilde X = \begin{pmatrix} X_1 & \cdots & X_G & X_{1:2} & \cdots & X_{1:G} & X_{2:3} & \cdots & X_{(G-1):G}  \end{pmatrix}
\end{equation}
where $X_{g:h}$ is the submatrix including $X_g$, $X_h$, and all
$p_gp_h$  column multiples between columns in group $g$ and columns
in group $h$. For example, if
\[
X_g = \begin{pmatrix} X_{g1} & X_{g2} \end{pmatrix}, \quad
X_h = \begin{pmatrix} X_{h1} & X_{h2} \end{pmatrix}
\]
are two groups each containing two columns, then
\[
X_{g:h} = \begin{pmatrix} X_g & X_h & X_{g1} * X_{h1} & X_{g1} * X_{h2} & X_{g1} * X_{h1} & X_{g2} * X_{h2} \end{pmatrix}
\]
where * denotes the pointwise product (Hadamard product) of vectors
(the $i$th entry of $X_{g1} * X_{h1}$ is the $i$th entry of $X_{g1}$ times
the $i$th entry of $X_{h1}$). Note that this expanded matrix has
$\tilde G = G + \binom{G}{2} = O(G^2)$ groups. We refer to the first
$G$ of these as main effects or main effect groups, and the remaining
as interaction groups. Finally, instead of fitting a
model by group lasso, we use forward stepwise on the expanded design
matrix. The overlapping groups still guarantee that our fitted model will
satisfy the strong hierarchy condition.

@@

\cite{HIV} use genomic information to predict efficacy of antiretroviral
drugs in treating HIV. Quantitative measurements of drug
response/resistance were regressed on categorical covariates encoded
the presence and type of certain genetic mutations. We attempt a similar
analysis using forward stepwise, and report our P-value at each step.
Categorical covariates are encoding as groups of binary variables using
the full encoding, and weights are set to the square-root of the
number of levels for that covariate. We perform forward stepwise once
for each drug response, restricting to the subset of the data with
no missing values and requiring all categorical variables to have 2 or
greater observations in each level.

@@

Results from two data sets are displayed in Figure~\ref{fig:HIVPI} and
Figure~\ref{fig:HIVNRTI}. Each panel shows results for a different drug,
with p-values plotted by step in forward stepwise. The solid black point
is the last p-value to fall under the 0.1 threshold and points after it
are colored red. If the \textit{last} stopping rule were used then the
p-values for including variables are the black points, and the resulting
models are shown in Tables 4 and 5. (\todo Fix this ref after re-running)

@@

Now fix $h$, let $P_g = X_gX_g^T$ and define
\begin{align*}
a_h &= w_h^{-1} X_h^T (I-P_g) y \\
b_h &= w_g w_h^{-1} X_h^T P_g y / \norm{X_g^T y} \\
c_h &= a_h^T b_h / (\norm{a_h} \norm{b_h}) \\
K_h   &= \{ x : \norm{x} = 1, b_h^Tx > w \}
\end{align*}
Note that $K_h = \emptyset$ if and only if $\norm{b_h} < w$. We will consider two cases. First, suppose $\norm{b_h} > w$. In this case we must rule out one sub-case, when $a_h$ is not in the polar cone of $K_h$. If this occurs, then on at least one side of the boundary of $K_h$ we have $a_h^Tx > 0$ there, so that the infimum inside $K$ is $-\infty$ but the supremum outside $K$ is $+\infty$. This violates the characterization of the global maximizer of the original process as described above, so this case cannot occur.

>>>>

Forward stepwise regression is a classical model selection procedure
that begins with an empty model and sequentially adds the best predictor
variable in each step. The usual $\chi^2$ and $F$-tests for significance
fail when a model has been selected this way and tend to be
anti-conservative unless they are computed on a held-out validation
dataset. In the lasso setting, \cite{significance:lasso} derived
a novel test statistic and its asymptotic null distribution, making
possible valid inferences after model selection using the full data.
\cite{tests:adaptive} modified and extended those results to the
group lasso \citep{grouplasso} and other penalized regression
problems, but only under the global null hypothesis.
One of the strengths of these test statistics is that they can be
used for valid significance testing when computed on the same
data used for model selection, eliminating the need for data splitting.
The present work iteratively applies the global null test
of \cite{tests:adaptive} for each step in forward stepwise selection,
and works out some of the details necessary for
models with grouped variables. The resulting method can
be more statistically efficient than validation on held-out data, and
more computationally efficient than penalized methods with
regularization parameters chosen by cross-validation.

@@

Some attempts to address this issue include Monte Carlo
estimation of tables of adjusted $F$-statistic values \citep{mc:ftoenter},
and permutation statistics \citep{permutation:stop}. Aside from the works
this paper is based on, there have been other recent attempts to do
inference after model selection. Most of these make use of subsampling
\citep{meinshausen:buhlmann} or data splitting \citep{wasserman:roeder}.
Our approach allows use of the full data and does not require the
extra computation involved in subsampling.
Before describing the full approach we first introduce notation and
specify our implementation of forward stepwise, which is slightly
different from the most commonly used versions.

We allow forward stepwise selection to add groups of variables in each
step, not only in the case of binary encoding for categorical
variables but also for any grouping purpose. For example, groups of
variables may be pre-designated factors such as expression
measurements for all genes in a single functional pathway.
To emphasize this we use $g, h$ as indices rather than the usual
$i, j$ throughout. Since single variables can be considered groups of
size 1, our general exposition includes non-grouped situations as a special
case. Our variant of forward stepwise uses a different method than
usual for choosing which group to add because we are using the
same objective in the forward stepwise procedure that was used to
derive the test statistics in penalized regression settings.

@@

Define $p = \sum_{g=1}^Gp_g$,
the total number of individual variables, so $p = G$ in the
case where all groups have size 1. Let $X$ be the matrix constructed
by column-binding the $X_g$, that is  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation*}
X = \begin{pmatrix} X_1 & X_2 & \cdots & X_G  \end{pmatrix}
\end{equation*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We assume throughout that the individual columns of $X$ are scaled to
have unit 2-norm (we do not scale groups, this is accomplished later
by weighting).
With each group we associate the $p_g \times 1$ coefficient vector
$\beta_g$, and write $\beta$ for the $p \times 1$ vector constructed
by stacking all of the $\beta_g$ in order.  Finally, our model for the
response is
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\begin{aligned}
\label{eq:gmodel}
y & = X \beta + \sigma \epsilon \\
   & = \sum_{g=1}^G X_g \beta_g + \sigma \epsilon
\end{aligned}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
where $\epsilon$ is noise. We assume Gaussian noise
$\epsilon \sim N(0, \Sigma)$ with known covariance matrix $\Sigma$.

The model \eqref{eq:gmodel} is underdetermined when $p > n$.
In such cases it still may be possible to
estimate $\beta$ well if it is sparse--that is, if it has few nonzero
entries. In the rest of this paper we refer to variable groups $X_g$
as noise groups if $\beta_g$ is a zero vector and as true or signal
groups if $\beta_g$ has any nonzero entries. We refer to the number
of such nonzero groups as the \textit{sparsity} of the model, and
denote this $k := \# \{ g : \beta_g \neq 0 \}$.

Before we describe the forward stepwise procedure we require one last
ingredient. To each group of variables we assign a weight $w_g$. These
weights act like penalties or costs, so increasing $w_g$ makes it
more difficult for the group $X_g$ to enter the model. The
modeler can choose weights arbitrarily, but we mostly use one
particular choice, based on $p_g$, that we discuss later. With this we
are ready to describe the forward stepwise procedure.

@@

\begin{algorithm}
  \caption{Forward stepwise variant with groups and weights}
  \label{algo:fs}
  \begin{algorithmic}[1]
    \REQUIRE An $n$ vector $y$ and $n \times p$ matrix $X$ of $G$ variable groups with weights $w_g$
    \ENSURE Ordered active set $A$ of variable groups included in the model at each step
    \STATE $A \gets \emptyset$, $A^c \gets \{ 1, \ldots, G\}$
    \FOR{$s=1$ to $steps$}
    \STATE $g^* \gets \argmax_{g \in A^c} \{ \norm{X_g^T r_{s-1}}_2 / w_g \}$
    \STATE $P_g \gets I_{n\times n} - X_{g^*}X_{g^*}^\dagger$
    \STATE $A \gets A \cup \{ g^* \}$, $A^c \gets A^c \backslash \{ g^* \}$
    \FORALL{$g \in A^c$}
      \STATE $X_g \gets P_g X_g$
    \ENDFOR
    \STATE $r_s \gets P_g r_{s-1}$
    \ENDFOR
    \RETURN $A$
  \end{algorithmic}
\end{algorithm}

The active set $A$ contains variable groups chosen to be included in
the model. Fitting $\hat \beta$ can be done by tracking the individual
fits and projections, or by simply fitting a linear model on the submatrix
of $X$ corresponding to $A$. 
We now use forward stepwise to refer specifically to
Algorithm~\ref{algo:fs} unless otherwise specified. Note that other
implementations of forward stepwise use different criteria for choosing
the next variable to add, such as the correlation with the residual.
Since we do not renormalize the columns after projecting the covariates
(lines 6 to 8 above), and since we have weights, we are not computing
correlations unless the design matrix is orthogonal and all weights are 1.
There are advantages and disadvantages to both criteria which we do not
discuss. Our choice was motivated by the group lasso result in
\cite{tests:adaptive}, but we believe other criteria can be handled
with appropriate modifications.

@@

\cite{significance:lasso} prove that under the null hypothesis that $A_k$
contains all the strong predictor variables, the distribution of a
test statistic $T_k \propto \lambda_k(\lambda_k - \lambda_{k+1})$ is
asymptotically Exp(1). In the lasso case we know a lot about the knots
and active set, but the group lasso picture is slightly more
complicated. For the group lasso, $\hat \beta(\lambda)$ does not have
piecewise linear components. To overcome this difficulty we
restrict our attention to the first group of variables to enter the
active set since the analysis then follows almost exactly as for the
lasso.

@@

We now describe how to calculate our p-value, following the discussion of
group lasso in \cite{tests:adaptive}. For the rest of this section let
$g = \gstar$ be the index of
the group attaining the maximum on line 3 of Algorithm~\ref{algo:fs}.
%\lambda = \norm{X_g^Ty}_2/w_g$ be the value of the maximum, and
%$r = \text{rank}(X_g)$ be the rank of the maximizing group.
For a vector $u \in \real^p$, let $u_h$ denote the coordinates
of $u$ corresponding to the columns of group $h$ in the design matrix $X$.
We can rearrange the columns of $X$ to group these adjacently, so that
\[
u^TX^T = \begin{pmatrix} u_1^T X_1^T & u_2^T X_2^T & \cdots & u_G^TX_G^T \end{pmatrix}
\]
One step of the calculation will be to find an orthonormal basis for the
linear space $\vecsp_g = \{ u \in \real^p : u_g^T X_g^T y = 0, u_h = 0
\text{ for all } h \neq g \}$ so that we can project orthogonally to
this space. If $X_g$ is a single column and $X_g^Ty \neq 0$ (which
should be the case since $g$ maximizes the absolute value of this
quantity), then the
space is trivial and the desired orthogonal projection is the identity.
Otherwise, if $X_g$ has $p_g > 1$ columns the space $\vecsp_g$
generally has dimension $p_g-1$.
% have dimension $r-1$ for $r = \text{rank}(X_g)$. This holds except in
% degenerate cases such as $X_g = 0$ or $r \ll p_g$, but these cases are not
% likely to occur since $\norm{X_g^T y}_2/w_g$ is supposed to be maximized.
% We proceed by forming the $n \times p_g$ matrix
% $X_g - X_g X_g^T y y^T X_g/\norm{X_g^Ty}_2^2$ and performing Gram-Schmidt
% on the first $r-1$ columns. For precise theorems showing that this will
% generally work, see \cite{outerp:indep}.
We compute an orthonormal basis by Gram-Schmidt and form a
$p_g \times p_g$ matrix, which we denote $V_g$, by appending one column
%0's in the coordinates corresponding to all groups $h \neq g$ and to
of zeroes (since Gram-Schmidt only produces $p_g-1$ columns).
With this we are ready to define the projection

@@

\begin{algorithm}
  \caption{Computing p-value}
  \label{algo:pval}
  \begin{algorithmic}
    \REQUIRE Response $y$, grouped design matrix $X$ with weights, inactive set $A^c$, index $g$ of last group to enter active set $A$.
    \ENSURE Our p-value for the group $g$ entering the model.
    \STATE Compute $H_g$ and $\sigma^2_g$
    \IF{$X_g$ is a single column}
    \STATE $\sigma_g^2 \gets X_g^T\Sigma X_g/w_g^2$
    \STATE $\tilde X_g \gets X_g / w_g \cdot \text{sign}(X_g^Ty)$
    \ELSE
    \STATE $P_g \gets \Sigma XV_g (V_g^T X^T \Sigma X V_g)^\dagger V_g^TX^T$
    \STATE $H_g \gets (I - P_g)\Sigma$
    \STATE $\sigma^2_g \gets y^TX_gX_g^T H_g X_gX_g^Ty / (w_g^2 \norm{X_g^Ty}_2^2)$
    \STATE $\tilde X_g \gets H_g X_g X_g^T y /(\norm{X_g^Ty}_2 w_g)$
    \ENDIF
    \STATE $\lambda \gets \norm{X_g^Ty}_2/w_g$
%    \STATE \COMMENT{Compute the following two $p$-vectors}
    \STATE $a \gets X^T(y - \lambda \tilde X_g)$
    \STATE $b \gets X^T \tilde X_g$
    \STATE Compute solution $(v_-, v_+)$ of \textsc{LinearFractional}(a,b) optimization subproblem
    \STATE $r \gets \text{rank}(X_g)$
    \STATE $u \gets [F_{\chi^2_r}(v^2_-/\sigma_g^2) - F_{\chi^2_r}(\lambda^2/\sigma_g^2)]/[F_{\chi^2_r}(v^2_-/\sigma_g^2) - F_{\chi^2_r}(v^2_+/\sigma_g^2)]$
    \RETURN $u$
  \end{algorithmic}
\end{algorithm}

We next describe the \textsc{LinearFractional} subproblem and its solution. We
gave the problem this name because it originated in the form
\[
\maximize_{h \neq g, \norm{u_h}_2 = 1} \frac{u_h^TX_h^Ty - u_h^TX_h^TX_gX_g^Ty}
{1 - u_g^TX_g^TX_hu_h}
\]
The solution we describe next is to a slightly different problem, and
also incorporates the information that $g$ maximizes $\norm{X_g^Ty}_2/w_g$.
For details see Appendix~\ref{app:closedform} or \cite{tests:adaptive}.

\begin{algorithm}
  \caption{The \textsc{LinearFractional} optimization subproblem}
  \label{algo:linfrac}
  \begin{algorithmic}
    \REQUIRE The $p$-vectors $a, b$ in Algorithm~\ref{algo:pval}, weights, inactive set $A^c$, a small tolerance number
%(we take $1e^{-10}$)
    \ENSURE Solution pair $(v_-, v_+)$.
    \FOR{$h$ in $A^c$}
      \IF{$\norm{b_h}_2 == 0$ or $\norm{a_h}_2/\norm{b_h}_2 < \text{tol}$}
        \STATE $(v^-_h, v^+_h) \gets (0, \infty)$
      \ELSE
        \STATE $\theta_c \gets a_h^Tb_h/(\norm{a_h}_2 \norm{b_h}_2)$
        \STATE $\theta_s \gets \sqrt{1-\theta_c^2}$
        \STATE $\theta \gets \arccos (\theta_c)$
        \STATE $\phi_s \gets \theta_s \norm{b_h}_2 / w_h$
        \IF{$\phi_s > 1$}
          \STATE $(v^-_h, v^+_h) \gets (0, \infty)$
        \ELSE
          \STATE $\phi \gets \arcsin(\phi_s)$
%          \STATE $\phi_2 \gets \pi - \phi_1$
          \STATE $z_\pm \gets s_\pm \norm{a_h}_2 \cos(\phi) /(w_h - s_\pm \norm{b_h}_2 \cos(\theta - \phi))$ for $s_\pm = \pm 1$
          \IF{$\norm{b_h}_2 < w_h$}
            \STATE $(v_h^-, v_h^+) \gets (\max \{ z_+, z_- \}, \infty)$
          \ELSE
            \STATE $(v_h^-, v_h^+) \gets (\min \{ z_+, z_- \}, \max \{ z_+, z_- \})$
          \ENDIF
        \ENDIF
      \ENDIF
    \ENDFOR
    \STATE $v^- \gets \min_h v_h^-$
    \STATE $v^+ \gets \max_h v_h^+$
    \RETURN $(v^-, v^+)$
  \end{algorithmic}
\end{algorithm}

@@

\subsection{Glinternet for hierarchical interactions}
\label{sec:glint}
In regression settings with many variables, considering models with pairwise interactions can drastically increase model complexity. \cite{glint} propose a method called \textsc{glinternet} to reduce the statistical complexity of this problem. The method imposes a strong hierarchical constraint on interactions (similar to that in \cite{bien:hierarchical}) where an interaction term can only be included if both its main effects are also included. They accomplish this by creating a design matrix with both main effects alone and also with groups including main effects with their first order interactions. Then they fit a group lasso model with the expanded design matrix. Because interaction terms only appear in groups along with their respective main effects, the hierarchy condition holds for the fitted model. We now consider a related procedure as an example problem, but first modify their method to simplify some parts. Let the expanded design matrix be given by
\begin{equation}
\label{eq:glintmat}
\tilde X = \begin{pmatrix} X_1 & \cdots & X_G & X_{1:2} & \cdots & X_{1:G} & X_{2:3} & \cdots & X_{(G-1):G}  \end{pmatrix}
\end{equation}
where $X_{g:h}$ is the submatrix including $X_g$, $X_h$, and all
$p_gp_h$  column multiples between columns in group $g$ and columns
in group $h$. For example, if
\[
X_g = \begin{pmatrix} X_{g1} & X_{g2} \end{pmatrix}, \quad
X_h = \begin{pmatrix} X_{h1} & X_{h2} \end{pmatrix}
\]
are two groups each containing two columns, then
\[
X_{g:h} = \begin{pmatrix} X_g & X_h & X_{g1} * X_{h1} & X_{g1} * X_{h2} & X_{g1} * X_{h1} & X_{g2} * X_{h2} \end{pmatrix}
\]
where * denotes the pointwise product (Hadamard product) of vectors
(the $i$th entry of $X_{g1} * X_{h1}$ is the $i$th entry of $X_{g1}$ times
the $i$th entry of $X_{h1}$). Note that this expanded matrix has
$\tilde G = G + \binom{G}{2} = O(G^2)$ groups. We refer to the first
$G$ of these as main effects or main effect groups, and the remaining
as interaction groups. Finally, instead of fitting a
model by group lasso, we use forward stepwise on the expanded design
matrix. The overlapping groups still guarantee that our fitted model
satisfies the strong hierarchy condition.

@@

\cite{HIV} use genomic information to predict efficacy of antiretroviral
drugs in treating HIV. Quantitative measurements of drug
response/resistance were regressed on categorical covariates encoded
the presence and type of certain genetic mutations. We attempt a similar
analysis using forward stepwise, and report our P-value at each step.
Categorical covariates are encoded as groups of binary variables using
the full encoding, and weights are set to the square-root of the
number of levels for that covariate. We perform forward stepwise once
for each drug response, restricting to the subset of the data with
no missing values and requiring all categorical variables to have 2 or
greater observations in each level.

@@

Results from two data sets are displayed in Figure~\ref{fig:HIVPI} and
Figure~\ref{fig:HIVNRTI}. The PI data contains protease inhibitor
resistance mutations, and the NRTI data contains nucleoside RT inhibitor
resistance mutations.
Each panel shows results for a different drug,
with p-values plotted by step in forward stepwise. The solid black point
is the last p-value to fall under the 0.1 threshold and points after it
are colored red. If the \textit{last} stopping rule were used then the
p-values for including variables are the black points, and the resulting
models are shown in Tables 4 and 5. (\todo Fix this ref after re-running)

@@

Now fix $h$, let $P_g = X_gX_g^T$ and define
\begin{align*}
a_h &= w_h^{-1} X_h^T (I-P_g) y \\
b_h &= w_g w_h^{-1} X_h^T P_g y / \norm{X_g^T y} \\
c_h &= a_h^T b_h / (\norm{a_h} \norm{b_h}) \\
K_h   &= \{ x : \norm{x} = 1, b_h^Tx > w \}
\end{align*}
Note that $K_h = \emptyset$ if and only if $\norm{b_h} < w$. We consider two cases. First, suppose $\norm{b_h} > w$. In this case we must rule out one sub-case, when $a_h$ is not in the polar cone of $K_h$. If this occurs, then on at least one side of the boundary of $K_h$ we have $a_h^Tx > 0$ there, so that the infimum inside $K$ is $-\infty$ but the supremum outside $K$ is $+\infty$. This violates the characterization of the global maximizer of the original process as described above, so this case cannot occur.
